---
title: Bayesian inference for state-space models with R
output: github_document
---

<!-- badges: start -->
![GitHub R package version](https://img.shields.io/github/r-package/v/epiverse-trace/bpmodels)
[![codecov](https://codecov.io/github/sbfnk/RBi/branch/master/graphs/badge.svg)](https://codecov.io/github/sbfnk/RBi) 
![GitHub contributors](https://img.shields.io/github/contributors/sbfnk/RBi)
 [![License: GPL v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

[rbi](https://github.com/sbfnk/RBi) is an `R` interface to [libbi](http://libbi.org), a library for Bayesian inference.

It mainly contains:
* various functions to retrieve and process the results from libbi (which are in NetCDF format)
* a `bi_model` class, to manipulate libbi models
* a `libbi` wrapper class, to perform Bayesian using libbi inference from within R,

Installation
==============

The easiest way to install the latest stable version of **rbi** is via [CRAN](https://cran.r-project.org/package=rbi):

```{r eval = FALSE}
install.packages("rbi")
```

Alternatively, the current development version can be installed using the `remotes` package

```{r eval = FALSE}
# install.packages("remotes")
library("remotes")
install_github("sbfnk/rbi")
```

The **rbi** package has only been tested on GNU/Linux and OS X, but it should mostly work everywhere `R` works.

If you want to use **rbi** as a wrapper to **LibBi** then you need a working version of [LibBi](https://github.com/lawmurray/LibBi). To install **LibBi** on a Mac, the easiest way is to install [Homebrew](https://brew.sh), followed by (using a command shell, i.e. Terminal or similar):

```{sh eval = FALSE}
brew install libbi
``` 

On linux, follow the [instructions](https://github.com/lawmurray/LibBi/blob/master/INSTALL_LINUX.md) provided with LibBi.

The path to `libbi` script can be passed as an argument to **rbi**, otherwise the package tries to find it automatically using the `which` linux/unix command.

If you just want to process the output from **LibBi**, then you do not need to have **LibBi** installed.

Getting started
==============

A good starting point is to look at the included demos:

```{r eval = FALSE}
 demo(PZ_generate_dataset) ## generating a data set from a model
 demo(PZ_PMMH)             ## particle Markov-chain Metropolis-Hastings
 demo(PZ_SMC2)             ## SMC^2
 demo(PZ_filtering)        ## filtering
```

For further information, have a look at the introductory vignette from the link from the [rbi CRAN package](https://CRAN.R-project.org/package=rbi).

Using coda
==========

**LibBi** contains the `get_traces` method which provides an interface to [coda](https://cran.r-project.org/package=coda).

Other packages
==============

For higher-level methods to interact with [LibBi](https://github.com/lawmurray/LibBi), have a look at [RBi.helpers](https://github.com/sbfnk/RBi.helpers).
